Handling Recursion in Generic Programming Using Closed Type Families

### 2018-08-20 lncs-paper.tex
Address remaining remarks and questions, cut the zipper

\lstset{
basicstyle=\ttfamily,
xleftmargin=\parindent,
aboveskip=0.2cm,
belowskip=0.2cm,
basewidth=0.5em,
numberbychapter=false,
escapeinside={(/}{/)},
literate={::}{{\ann}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\ds}}2
{>>>}{{\comp}}2 {>=>}{{\kleisli}}2
}

@@

\newcommand{\nhs}{\hspace{-0.06cm}}

@@

The problem of using overlapping instances was first addressed by Kiselyov~et~al.~\cite{Kiselyov2004}. Their technique for avoiding overlap relies on a Haskell 98 extension for functional dependencies. The solution proposes two variants of defining a type-level equality predicate, a type class, and then systematically localizes overlap by circumventing it with that predicate. The first version of type equality maps types to unique type representations and compares them. Its later implementation with type families~\cite{Kiselyov2012} fully eliminates \textsf{OverlappingInstances}. Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC. The most generic solution for type equality, the second version, again makes use of overlapping instances, however.

@@

\subsection{Problem with handling recursion}
\label{sec:recursion-problem}

@@

The SOP-based \K{subterms} definition is outlined below:
\begin{lstlisting}
subterms :: Generic a => a -> [a]
subterms t = subtermsNS $ unSOP $ from t

@@

Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type. There is no clear path to this at the moment. Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
\begin{lstlisting}
class Subterms a (xs :: [*]) where
  subtermsNP :: NP I xs -> [a]

@@

\subsection{Solution to \K{subtermsNP} revised}
\label{subsec:solve-subtermnp}

We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
\begin{lstlisting}
class DecideEq (eq :: Bool) (a :: *) (b :: *) where
  decideEq :: Maybe (b :==: a)

@@

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
\begin{lstlisting}
class    DecideEq (Equal a b) a b => ProofEq a b
instance DecideEq (Equal a b) a b => ProofEq a b
\end{lstlisting}
The equality proof can then be employed to provide a type-safe cast between two equal types:
\begin{lstlisting}
castEq :: ∀a b. ProofEq a b => b -> Maybe a
castEq t = (\d -> castWith d t) <$> decideEq @(Equal a b)
\end{lstlisting}
Resulting implementation of \K{subtermsNP} resembles our first definition given in the previous section:
\begin{lstlisting}
subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
subtermsNP (I (y :: x) :* ys)
  = case castEq y of
      Just t  -> t : subtermsNP ys
      Nothing -> subtermsNP  ys
subtermsNP _ Nil = []
\end{lstlisting}

@@

To complete the solution of the problem, the \K{ProofEq} constraint must be added to the \K{subterms} and \K{subtermsNS} declarations as well.

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \K{subterms}' task. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

@@

The function \K{show} is a common example of useful functions that traverse a datatype's recursive structure. It is known that this function can be defined in a generic way for an arbitrary datatype. De Vries and L\"{o}h define generic function \K{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view. We follow their implementation of \K{gshow} for the most part, but improve it in respect of handling recursion. The original \K{gshow} yields to the standard \K{show} generated through \K{deriving Show}, because it does not consult with recursion points to place parentheses. We eliminate this drawback.

@@

instance Show  b => CaseShow 'False a b where
  caseShow'   = show
instance GShow a => CaseShow 'True  a a where
  caseShow' t = "(" ++ gshow t ++ ")"
\end{lstlisting}
We provide a synonym for the \K{CaseShow (Equal a b) a b} instance, which we call \K{CaseRecShow}, as before with \K{ProofEq}; likewise a synonym for the matching function:
\begin{lstlisting}
caseShow :: ∀a b. CaseRecShow a b => b -> String
caseShow t = caseShow' @(Equal a b) @a t
\end{lstlisting}

@@

\section{Polymorphic recursion}
\label{sec:polymorphic-recursion}

The solution, described in the previous section, is applicable to a range of datatypes that are \emph{monomorphically recursive}. Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g. \K{Tree a} from Section~\ref{sec:sop-view}). We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent. It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested datatypes.

The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}. The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
\begin{lstlisting}
instance Show  a => CaseShow 'False a where
  caseShow'   = show
instance GShow a => CaseShow 'True  a where
  caseShow' t = "(" ++ gshow t ++ ")"
\end{lstlisting}

Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
\begin{lstlisting}
data Nested a = Epsilon | Nest a (Nested [a])
\end{lstlisting}
The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow b} instance context above.

@@

In this section, we describe our approach allowing one to define the generic zipper out of a representation that does not exploit a fixed point. We start with the generic zipper interface and an example of how it can be used (Section~\ref{ss:zip-interface}). Then, we turn to the type-level machinery employed to define locations inside mutually recursive datatypes using the SOP view  (Section~\ref{ss:locs}). Finally, we discuss the implementation of the generic zipper interface~--- the functions for manipulating locations (Section~\ref{ss:zip-impl}).

@@

Our zipper applies to regular datatypes as well. In that case, \K{fam} list shall contain a single element. Generally, the interface is flexible enough to allow us to check in any collection of types we are interested in during traversal. However, we demand an updating operation to be a type class function to distinguish the types of the nodes.

@@

The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
\begin{lstlisting}
data Loc (r :: *) (fam :: [*]) (c :: * -> Constraint) where
  Loc :: Focus r a fam c -> Contexts r a fam c
      -> Loc r fam c
\end{lstlisting}
The type parameters \K{r}, \K{fam}, and \K{c} in \K{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively. Also, the single constructor is existentially quantified over one more type variable, \K{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure. We discuss both term parameters of the constructor of \K{Loc} in detail below.

@@

The subtree in focus is wrapped by the \K{Focus} datatype. 
The wrapper incapsulates the proofs about a number of important properties of a focus.
\begin{lstlisting}
data Focus (r :: *) (a :: *) (fam :: [*])
           (c :: * -> Constraint) where
 Focus :: (Generic b, In b fam, ZipperI r a b fam c)
       => b -> Focus r a fam c
\end{lstlisting}
Existential type variable \K{b} represents the type of a focus. We apply a number of predicates to \K{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus. Firstly, the type of a focus should have the \K{Generic} representation. Secondly, it should live \K{In} the list of types we are going to visit. Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \K{ZipperI} predicate. In particular, the predicate ensures that \K{a} is the type of the parent for the focus in the structure under consideration.

@@

One last missing piece for managing focuses is the class \K{ProofIn}. It provides a proof of membership of a focus type to a family. Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}. The definitions of \K{ProofIn} and an auxiliary class \K{ProofFocus} are given in Figure~\ref{fig:proof-in}. In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak form of proof, which is more flexible.
%The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
% this is not clear and, I think, can be omitted

\begin{figure}[t]
\begin{CenteredBox}
{\lset
\begin{lstlisting}
class ProofFocus (inFam :: Bool) (r :: *) (a :: *) (b :: *)
                 (fam :: [*]) (c :: * -> Constraint) where
  witness :: b -> Maybe (Focus r a fam c)

instance ProofFocus 'False r a b fam c where
  witness _ = Nothing
instance (Generic b, In b fam, ZipperI r a b fam c)
    => ProofFocus 'True r a b fam c where
  witness   = Just . Focus

class ProofFocus (InFam b fam) r a b fam c
    => ProofIn r a b fam c
instance ProofFocus (InFam b fam) r a b fam c
    => ProofIn r a b fam c
\end{lstlisting}}
\end{CenteredBox}
\caption{Proof of membership of a family of datatypes.}
\label{fig:proof-in}
\end{figure}

@@

McBride~\cite{McBride2001} studies a relation between the one-hole context definition and \emph{partial differentiation} from calculus: he shows that the type of the context for an arbitrary (regular) type can be derived mechanically from that type by means of a list of differentiation \emph{rules} that serve as formulaic instructions for computing the type in type-level programming. Yakushev~et~al.~\cite{MuRec2009} then demonstrate that the method can be generalized for mutually recursive datatypes. We adapt that technique to \textsf{generics-sop}, and now need a few auxiliary type-level functions to implement the computation of the context type. Those functions, defined recursively via type families, provide algebraic operations for lists of types (which we regard as sums and products of types): addition and multiplication. Specifically, we define addition \K{.++} of two sums of products (SOP) of types, multiplication \K{.*} of a SOP by a single type, and multiplication \K{.**} of a SOP by a product of types, as shown in Figure~\ref{fig:type-arithmetic}.
\begin{figure}[h]
\begin{CenteredBox}
{\lset
\begin{lstlisting}
(/\textbf{\textrm{SOP addition}}/)

type family (.++) (xs :: [[*]]) (ys :: [[*]]) :: [[*]] where
  (x ': xs) .++ ys = x ': (xs .++ ys)
  '[]       .++ ys = ys
infixr 6 .++

(/\vspace{-0.3cm}/)
(/\textbf{\textrm{SOP-by-type multiplication}}/)

type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
  x .* (ys ': yss) = (x ': ys) ': (x .* yss)
  _ .* '[]         = '[]
infixr 7 .*

(/\vspace{-0.3cm}/)
(/\textbf{\textrm{SOP-by-product multiplication}}/)

type family (.**) (xs :: [*]) (ys :: [[*]]) :: [[*]] where
  (x ': xs) .** yss = x .* (xs .** yss)
  '[]       .** yss = yss
infixr 7 .**
\end{lstlisting}}
\end{CenteredBox}
\caption{Algebraic operations on type-level sums and products.}
\label{fig:type-arithmetic}
\end{figure}

The addition operation just appends two type-level lists of lists (sums of products), multiplication by a type adds the type to the head of each inner list of the sum (here we see multiplication of a product and the distributive property of multiplication over addition, just as in arithmetic of numbers), and multiplication by a product appends the list to the head of each inner product of the sum. Again, kind \K{[*]} here denotes products, and \K{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers in these definitions becomes more clear if one realizes that an empty sum \K{'[] :: [[*]]} corresponds to $1$, and an empty product \K{'[] :: [*]} corresponds to $0$. We also specify, through the \K{infixr} declaration, that multiplication has a higher priority than addition.

@@

At this point, we can implement differentiation of a product of types and, therefore, the computation of a context frame type.

The definition of differentiation, shown in Figure~\ref{fig:diff-prod}, resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \K{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$. When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \K{Hole}. Reflecting this type in the context is helpful when we traverse the context representation to plug the hole. We use the empty type \K{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice. The sum \K{'[ '[]]} represents a unit type that is exactly $1$. We also use type-level \K{If} that returns its second argument for \K{'True}, and the third one otherwise. We do not give its definition here, as it is straightforward.
\begin{figure}[t]
\begin{CenteredBox}
{\lset
\begin{lstlisting}
data Hole = Hole
data End

type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
  DiffProd _   '[]       = '[]
  DiffProd fam '[x]      = If (InFam x fam) '[ '[Hole]] '[]
  DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]     '[]
  DiffProd fam (x ': xs)
    = Hole .* xs .** DiffProd fam '[End, x] .++
              x  .*  DiffProd fam xs
\end{lstlisting}}
\end{CenteredBox}
\caption{Differentiation of a product of types.}
\label{fig:diff-prod}
\end{figure}

The following completes the computation of the context type:
\begin{lstlisting}
data ConsN = F | N ConsN | None
  deriving Eq

type family ToContext (n :: ConsN) (fam :: [*])
                      (code :: [[*]]) :: [[*]] where
  ToContext _ _ '[] = '[]
  ToContext n fam (xs ': xss)
    = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
\end{lstlisting}
The type family \K{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code. Since each product from the code matches a sum of multiple products in the context, it is helpful for each product of the context representation, to keep the index of its matching constructor of the datatype. We store the index in the datatype \K{ConsN} adding that to the head of each product but wrapped by \K{Proxy} because we use \K{ConsN} promoted to a kind in the type family, while the products contain types of kind \K{*}. The constructors \K{F} and \K{N} denote ``first'' and ``next'', respectively, and the special constructor \K{None} will be used further to indicate failure of matching indices.

Finally, a context frame has two representations: as a type synonym and as a newtype wrapper:
\begin{lstlisting}
type    CtxCode fam a = ToContext 'F fam (Code a)

newtype Context fam a = Ctx {ctx :: SOP I (CtxCode fam a)}
\end{lstlisting}
The wrapper allows GHC to perform type inference where it is doomed to fail with a plain type synonym. On the other hand, the \K{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.

\subsection{Implementing the zipper interface}
\label{ss:zip-impl}

We now can implement the interface functions of the zipper, which we have previously described. We only demonstrate the implementation of the \K{goDown} function here. This shows the idea of how we can use our idiom for defining the zipper functions, and the source code with the full implementation of the zipper interface is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.

To move focus down to the leftmost child of the current focal node in the tree, we should analyze the focal subtree's representation to find its first immediate child, and compute its respective context. The following definition of \K{goDown} uses two auxiliary functions: \K{toFirst} and \K{toFirstCtx}.
\begin{lstlisting}
goDown :: Loc a fam c -> Maybe (Loc a fam c)
goDown (Loc (Focus t) cs)
  = case toFirst t of
      Just t' -> Just $ Loc t' (Ctxs (toFirstCtx t) cs)
      _       -> Nothing
\end{lstlisting}
The function \K{toFirst} returns its result in the \K{Maybe} monad, and may return \K{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down. The function \K{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.

\begin{figure}[t]
\begin{CenteredBox}
{\lset
\begin{lstlisting}
toFirst :: ∀fam c r a. (Generic a, ToFirst r a fam c)
        => a -> Maybe (Focus r a fam c)
toFirst t = appToNP @AllProof toFirstNP $ unSOP $ from t

(/\vspace{-0.3cm}/)
(/\textbf{\textrm{Proof}}/)

class    All (ProofIn r a fam c) xs => AllProof r a fam c xs
instance All (ProofIn r a fam c) xs => AllProof r a fam c xs

type ToFirst r a fam c = All (AllProof r a fam c) (Code a)

(/\vspace{-0.3cm}/)
(/\textbf{\textrm{Processing products}}/)

>>>>

\lstset{
basicstyle=\ttfamily,
xleftmargin=\parindent,
aboveskip=0.2cm,
belowskip=0.2cm,
basewidth=0.5em,
numberbychapter=false,
escapeinside={(/}{/)},
literate={::}{{\ann}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\ds}}2
{>>>}{{\comp}}2 {>=>}{{\kleisli}}2
{_1}{{$_\mathtt{1}$}}1 {_2}{{$_\mathtt{2}$}}1
{_n}{{$_\mathtt{n}$}}1 {_k}{{$_\mathtt{k}$}}1
{_Aux}{{\large$_\mathtt{\hs Aux}$}}2
{__Aux}{{\large$_\mathtt{Aux}$}}2
}

@@

\newcommand{\hs}{\hspace{0.1cm}}
\newcommand{\nhs}{\hspace{-0.06cm}}

@@

The problem of using overlapping instances was first addressed by Kiselyov~et~al.~\cite{Kiselyov2004}. Their technique for avoiding overlap relies on a Haskell 98 extension for functional dependencies. The solution proposes two variants of defining a type-level equality predicate, a type class, and then systematically localizes overlap by circumventing it with that predicate. The first version of type equality maps types to unique type representations and compares them. That variant and its later implementation with type families~\cite{Kiselyov2012} fully eliminate \textsf{OverlappingInstances}. Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC. The most generic solution for type equality, the second version, again makes use of overlapping instances, however.

@@

\subsection{Problem with handling recursion}
\label{subsec:recursion-problem}

@@

We describe a straghtforward implementation based on plain pattern matching, for reasons of simplicity. However, in Section~\ref{subsec:solve-subterms}, we show a better high-level solution using one of the primary advantages of \textsf{generics-sop}---high-level traversal functions.

The SOP-based \K{subterms} definition is outlined below:
\begin{lstlisting}
subterms :: Generic a => a -> [a]
subterms = subtermsNS . unSOP . from

@@

Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type. There is no clear path to this at the moment. Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
\begin{lstlisting}
class Subterms a (xs :: [*]) where
  subtermsNP :: NP I xs -> [a]

@@

\subsection{Solution to \K{subterms} revised}
\label{subsec:solve-subterms}

We now return to our running example from Section~\ref{subsec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
\begin{lstlisting}
class DecideEq (eq :: Bool) (a :: *) (b :: *) where
  decideEq :: Maybe (b :==: a)

@@

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \K{All} combinator and partially applied dedicated type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
\begin{lstlisting}
class    DecideEq (Equal a b) a b => ProofEq a b
instance DecideEq (Equal a b) a b => ProofEq a b
\end{lstlisting}
The equality proof can then be employed to provide a type-safe cast between two equal types:
\begin{lstlisting}
castEq :: ∀a b. ProofEq a b => b -> Maybe a
castEq t = (\d -> castWith d t) <$> decideEq @(Equal a b)
\end{lstlisting}

Below we show two variants of completing the \K{subterms} implementation. The low-level version performs pattern matching on the structure of \K{NS} and \K{NP}, as we do in Section~\ref{subsec:recursion-problem}. For this variant, we reimplement \K{subtermsNP}. The different implementation employes powerful machinery of high-level combinators---one of the major advantages provided by \textsf{generics-sop}. Although, those high-level functions become one extra layer of complexity.

\subsubsection{Low-level implementation}

The low-level definition of \K{subtermsNP} using the type cast resembles its outline given in the previous section:
\begin{lstlisting}
subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
subtermsNP (I (y :: x) :* ys) =
  case castEq y of
    Just t  -> t : subtermsNP ys
    Nothing -> subtermsNP  ys
subtermsNP _ Nil = []
\end{lstlisting}

@@

To finish up, the \K{ProofEq} constraint must be added to the \K{subterms} and \K{subtermsNS} declarations as well.

\subsubsection{High-level implementation}

By taking advantage of \textsf{generics-sop}'s powerful functions for collapsing and mapping, one can define the functions \K{subtermsNS} and \K{subtermsNP} as follows:
\begin{lstlisting}
subtermsNS :: ∀a xss. All (AllProofEq a) xss
           => NS (NP I) xss -> [a]
subtermsNS = collapse_NS
           . cmap_NS (Proxy @(AllProofEq a)) subtermsNP

subtermsNP :: ∀a xs. AllProofEq a xs => NP I xs -> K [a] xs
subtermsNP = K . catMaybes
               . collapse_NP
               . cmap_NP (Proxy @(ProofEq a)) (mapIK $ castEq)
\end{lstlisting}
This implementation requires also defining \K{AllProofEq} and properly changing the \K{subterms} type annotation.

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for the task of \K{subterms}. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

@@

The function \K{show} is a common example of useful functions that traverse a datatype's recursive structure. It is known that this function can be defined in a generic way for an arbitrary datatype. De Vries and L\"{o}h define generic function \K{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view. We follow their implementation of \K{gshow} for the most part, but improve it in respect of handling recursion. The example shows, how the better implementation, consulting with recursive positions, affects the usability of the function, obviating boilerplate code.

@@

instance Show  b => CaseShow 'False a b where
  caseShow' = show
instance GShow a => CaseShow 'True  a a where
  caseShow' = gshow
\end{lstlisting}
We provide a synonym for the \K{CaseShow (Equal a b) a b} instance, which we call \K{CaseRecShow}, as before with \K{ProofEq}; likewise a synonym for the matching function:
\begin{lstlisting}
caseShow :: ∀a b. CaseRecShow a b => b -> String
caseShow t = caseShow' @(Equal a b) @a t
\end{lstlisting}

@@

\subsection{Abstract description of the design pattern}
\label{subsec:abstract-pattern}

We can continue along the lines of the previous examples to consider the approach separately from concrete ones. One can use the technique abstractedly, following the pattern below:

\begin{lstlisting}
class DispatchRec_Aux (p :: Bool)
                    (a_1 :: k_1) ... (a_n :: k_n) (b :: *) where
  dispatchRec_Aux :: b -> X

instance C_1 ... => DispatchRec_Aux 'False a_1 ... a_n b where
  dispatchRec_Aux = f_1
instance C_2 ... => DispatchRec_Aux 'True  a_1 ... a_n b where
  dispatchRec_Aux = f_2

class    DispatchRec_Aux (P a_1 ... a_k b) a_1 ... a_k ... a_n b
  => DispatchRec a_1 ... a_n b
instance DispatchRec_Aux (P a_1 ... a_k b) a_1 ... a_k ... a_n b
  => DispatchRec a_1 ... a_n b

dispatchRec :: ∀a_1 ... a_n b. DispatchRec a_1 ... a_n b => b -> X
dispatchRec = dispatchRec_Aux @(P a_1 ... a_k b) @a_1 ... @a_k ... @a_n
\end{lstlisting}

This machinery employes the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec__Aux}. The predicate \K{P} generalizes \K{Equal} to an arbitrary relation on types, so that one can model, e.g., subtyping for some domain-specific language. The introduction of the associated type class \K{DispatchRec} provides a handy interface for dispatch. The dispatch function \K{dispatchRec} decides on one of the functions \K{f_1} and \K{f_2}---depending on whether the predicate \K{P} is true for types \K{a_1}, \dots, \K{a_k}, \K{b}.

@@

In this section, we describe our approach allowing one to define the generic zipper out of a representation that does not exploit a fixed point. We start with the generic zipper interface and an example of how it can be used (Section~\ref{ss:zip-interface}). Then, we turn to the type-level machinery employed to define locations inside mutually recursive datatypes using the SOP view  (Section~\ref{ss:locs}).

In Technical Report, we discuss the implementation of the generic zipper interface---the functions for manipulating locations. The source code with the full implementation of the zipper is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.

@@

Our zipper applies to regular datatypes as well. In that case, \K{fam} list shall contain a single element. Generally, the interface is flexible enough to allow us to check in any collection of types we are interested in during traversal. However, we demand an updating operation to rely on a type class function to distinguish the types of the nodes.

@@

The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
\begin{lstlisting}
data Loc (r :: *) (fam :: [*]) (c :: * -> Constraint) where
  Loc :: Focus r a fam c -> Contexts r a fam c
      -> Loc r fam c
\end{lstlisting}
The type parameters \K{r}, \K{fam}, and \K{c} in \K{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively. Also, the single constructor is existentially quantified over one more type variable, \K{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure. We discuss both the term parameters of the constructor of \K{Loc} in detail below.

@@

The subtree in focus is wrapped by the \K{Focus} datatype. 
The wrapper incapsulates the proofs about a number of important properties of a focus.
\begin{lstlisting}
data Focus (r :: *) (a :: *) (fam :: [*])
           (c :: * -> Constraint) where
 Focus :: (Generic b, In b fam, ZipperI r a b fam c)
       => b -> Focus r a fam c
\end{lstlisting}
Existential type variable \K{b} represents the type of a focus. We apply a number of predicates to \K{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus. Firstly, the type of a focus should have the \K{Generic} representation. Secondly, it should live \K{In} the list of types we are going to visit. Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \K{ZipperI} predicate. In particular, the predicate ensures that \K{a} is the type of the parent for the focus in the structure under consideration. Also, it guarantees that \K{b} fulfils the constraint \K{c}. Again, we make \K{c} embedded in \K{Loc} by reason of existentiality of \K{b}, which precludes the enforcement of this restriction by means of external \K{All}.

@@

One last missing piece for managing focuses is the class \K{ProofFocus}. It provides a proof of membership of a focus type to a family. Again, this generalizes (in a weak form) the proof of type equality from Section~\ref{subsec:solve-subterms}. The definitions of \K{ProofFocus} and an auxiliary class \K{ProofFocus__Aux} are given in Figure~\ref{fig:proof-in}. In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak kind of proof, which is more flexible. We can then make it handy by defining a function \K{castFocus} using \K{castFocus__Aux}, as with \K{castEq} before.
%The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
% this is not clear and, I think, can be omitted

\begin{figure}[t]
\begin{CenteredBox}
{\lset
\begin{lstlisting}
class ProofFocus_Aux (inFam :: Bool) (r :: *) (a :: *) (b :: *)
                   (fam :: [*]) (c :: * -> Constraint) where
  castFocus_Aux :: b -> Maybe (Focus r a fam c)

instance ProofFocus_Aux 'False r a b fam c where
  castFocus_Aux _ = Nothing
instance (Generic b, In b fam, ZipperI r a b fam c)
      => ProofFocus_Aux 'True  r a b fam c where
  castFocus_Aux   = Just . Focus

class    ProofFocus_Aux (InFam b fam) r a b fam c
    => ProofFocus r a b fam c
instance ProofFocus_Aux (InFam b fam) r a b fam c
    => ProofFocus r a b fam c
\end{lstlisting}}
\end{CenteredBox}
\caption{Proof of membership of a family of datatypes.}
\label{fig:proof-in}
\end{figure}

@@

McBride~\cite{McBride2001} studies a relation between the one-hole context definition and \emph{partial differentiation} from calculus: he shows that the type of the context for an arbitrary (regular) type can be derived mechanically from that type by means of a list of differentiation \emph{rules} that serve as formulaic instructions for computing the type in type-level programming. Yakushev~et~al.~\cite{MuRec2009} then demonstrate that the method can be generalized for mutually recursive datatypes. We adapt that technique to \textsf{generics-sop}, and now need a few auxiliary type-level functions to implement the computation of the context type. Those functions, defined recursively via type families, provide algebraic operations for lists of types (which we regard as sums and products of types): addition and multiplication. Specifically, we introduce addition \K{.++} of two sums of products (SOP) of types, multiplication \K{.*} of a SOP by a single type, and multiplication \K{.**} of a SOP by a product of types. The definition of the latter uses \K{.*}.

The addition operation just appends two type-level lists of lists (sums of products), multiplication by a type adds the type to the head of each inner list of the sum (here we see multiplication of a product and the distributive property of multiplication over addition, just as in arithmetic of numbers), and multiplication by a product appends the list to the head of each inner product of the sum. We define multiplication by a type below. The definitions for the other type-level operations are similar.
\begin{lstlisting}
type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
  x .* (ys ': yss) = (x ': ys) ': (x .* yss)
  _ .* '[]         = '[]
\end{lstlisting}

Again, kind \K{[*]} denotes products, and \K{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers becomes more clear if one realizes that an empty sum \K{'[] :: [[*]]} corresponds to $0$, and an empty product \K{'[] :: [*]} corresponds to $1$.

@@

At this point, we can implement differentiation of a product of types and, therefore, the computation of a context frame type\footnote{The actual definitions slightly differ from ones, presented in the following. We omit some implementation details for simplicity.}.

The definition of differentiation resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes:
\begin{lstlisting}
type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
  DiffProd _   '[]       = '[]
  DiffProd fam '[x]      = If (InFam x fam) '[ '[]] '[]
  DiffProd fam (x ': xs) =
    xs .** DiffProd fam '[x] .++ '[x] .** DiffProd fam xs
\end{lstlisting}
The differentiation of the single type reflected by a one-element list here results in $0$ reflected by \K{'[]}, if that is not in the family and hence is regarded as a constant. Otherwise it results in $1$ represented by the sum \K{'[ '[]]}. When differentiation gives $1$, it is actually the hole. We also use type-level \K{If} that returns its second argument for \K{'True}, and the third one otherwise. We do not give its definition here, as it is straightforward.

The following completes the computation of the context type:
\begin{lstlisting}
type family ToContext (fam :: [*])
                      (code :: [[*]]) :: [[*]] where
  ToContext _   '[]         = '[]
  ToContext fam (xs ': xss) =
    DiffProd fam xs .++ ToContext fam xss
\end{lstlisting}
The type family \K{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code.

Finally, the type of a context frame representation is a type synonym:
\begin{lstlisting}
type Context fam a = SOP I (ToContext fam (Code a))
\end{lstlisting}

\section{Discussion}
\label{sec:discussion}

In this section, we give a discussion of some concerns relating to the scope and user-friendliness of the introduced no-overlap technique for handling recursion.

\subsubsection{Polymorphic recursion}

The approach, described in this paper, is applicable to a range of datatypes that are \emph{monomorphically recursive}. Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g. \K{Tree a} from Section~\ref{sec:sop-view}). We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent. It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested~\cite{Bird1998} datatypes.

The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}. The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
\begin{lstlisting}
instance Show  a => CaseShow 'False a where
  caseShow' = show
instance GShow a => CaseShow 'True  a where
  caseShow' = gshow
\end{lstlisting}

Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
\begin{lstlisting}
data Nested a = Epsilon | Nest a (Nested [a])
\end{lstlisting}
The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow a} instance context above.

Nevertheless, there is a useful example of a generic operation that uses the recursive structure of data, but has not recursive calls in itself---and, therefore, can deal with nested datatypes. \emph{Generic generic programming}~\cite{MagLoeh2014} is a technique of defining translations between different generic programming universes. In our setting, it is possible to define a translation from \textsf{generics-sop} to another library with an explicit encoding of recursive positions, such as \textsf{regular}~\cite{VanNoort2008}. This usually needs only to detect recursion points, because the most of generic views provide shallow conversion functions.

\subsubsection{Usability}

The shown no-overlap technique may involve a lot of multiple constraints on generic functions. Those complex constraints introduce certain limitations to the approach and make it difficult in use. Especially, making a mistake in type-level programming with constraints which can usually be abbreviated in a code, one can get an embarrassing long error message where multiple constraints are displayed. Unfortunately, compound error messages on deducing constraints complicate the use of the technique, particularly for non-specialists in datatype-generic programming, in comparison with the existing approaches to treating recursion.
